HO unification from object language to meta language

### 2024-02-11 main.tex
Update tex

\def\elpi{\proglang{elpi}}
\def\coqelpi{\proglang{coq-elpi}}
\def\lambdaprolog{\proglang{$\lambda$-prolog}}
\def\coq{\proglang{coq}}
\newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}

@@

Meta programming \cite{hill1994} is a programming technique in which a program
can treat an other program as its data. This latter program is called object
language, while the former is called meta language. At the heart of meta
programming lays the necessity of representing terms of the object language in
the meta language so that a wide set of program manipulation ranging from
interpretation to compilation. Furthermore, since a logic program is a set of
clauses, the behavior of the object language can be modified by simply adding
new rules inside the database of the meta program.

The motivation pushing us to reason about unification of terms of the object
language represented as terms of the meta language is to implement a type class
solver for the \coq proof assistant into the \elpi logic programming language. A
type class \cite{wadler89, sozeau08} is a typing structure allowing to introduce
ad hoc polymorphism in functional languages and instances are implementations of
type classes. From the point of view of logic programming, type class represent
predicates parametrized by their argument and instances rules for those
predicates. For instance, if \texttt{XXX} is the type class representing the
AAA, then ZZZ and WWW are instances for XXX. In the code snippet below, we give
such implementation in \coq.

Type-class resolution starts from a query, that is a class applied to some
argument. This \coq term is translated into a term of the meta language and the
search for a solution in the database is started. However, since unification is
performed on the head of the clause, it may happen that the term representation
in the meta language may hide some unification properties that are true in the
object language. In the example below, the goal \texttt{Decision (@eq T a b)}
for some \texttt{a} and \texttt{b} unifies with \texttt{Decision (R x y)} in the
object language but not in its meta representation. Similarly, the goal
\texttt{RelDecision (@eq (sig T ?P))} where \texttt{?P}, under the hypothesis
\texttt{RelDecision (@eq nat)}, will trying to apply the rule for
\texttt{sig\_eq\_dec}, we fall into an higher order unification problem, where
\texttt{P} is applied to the local name \texttt{x}. However, the corresponding
rule in the meta language exploit a first order variable \texttt{P}. Therefore,
after the refinement of the goal to \texttt{sig\_eq\_dec}, the resolution
immediately fail to solve the premise \texttt{tc-ProofIrrel (app [P, c0]) (app
[P1, c0])}.

@@

In \coq, terms unify up to $\eta\beta\delta\iota\zeta$-equivalence and
unification is also possible when dealing with higher order variables. Whereas
\elpi allows higher order unification up to $\eta\beta$-equivalence, and the
representation of \coq terms is more syntactical than one would expect, making 
two terms that unify in \coq not unifiable in \elpi.

Without any loss of generality, in \cref{table:coq2elpi} we present a subset of
the terms in the object language and for each term we give its representation in
the meta language. This encoding is enough to fall into unification problems.
For example, rules like

\begin{center}
  \elpiIn{p (lam x\ (app F x)) :- ... F ...}
\end{center}

cannot unify with goals like

\begin{center}
  \elpiIn{p (lam x\ app (app (const "f") x) x).}
\end{center}

Since \elpiIn{F} does not see \elpiIn{x} even though it is applied to \elpiIn{x}
in the object language.

\begin{table}
  \centering
  \begin{tabular}{c||c|c}
  Term        & Obj. lang.                    & Meta lang.                                   \\
  \hline\hline
  Constant    & a                             & `a'                                     \\
  Application & f a\_1 a\_2 ... a\_n          & app{[}`f', `a\_1', `a\_2', ..., `a\_n'{]}     \\
  Abstraction & fun (x : T) $\Rightarrow$ f x & fun `x' T (x\textbackslash app{[}`f', `x'{]}) \\
  Variable    & X                             & X                                             \\
  \end{tabular}
  \caption{Obj. lang. terms to meta lang. terms representation}
  \label{table:coq2elpi}
\end{table}

\subsection{First-order unification}

@@

>>>>

\def\elpi{\proglang{elpi}}
\def\coqelpi{\proglang{coq-elpi}}
\def\lambdaprolog{\proglang{$\lambda$-prolog}}
\def\coq{\proglang{coq}}

\newcommand*{\acronym}[1]{\texttt{#1}\xspace}

\def\ol{\acronym{ol}}
\def\ml{\acronym{ml}}

\newcommand*{\eqtau}{\ensuremath{\mathrel{\overset{\mathrm{\tau}}{=}}}}

@@

Meta programming \cite{hill1994} is a programming technique in which a program
can treat an other program as its data. This latter program is called object
language (\ol for short), while the former is called meta language (\ml for
short). At the heart of meta programming lays the necessity of representing
terms of the \ol in the \ml so that a wide set of program
manipulations ranging from interpretation to compilation. Furthermore, since a
logic program is a set of clauses, the behavior of the \ol can be
modified \texttt{in corsa}\todo{translate in english} by simply adding new rules inside the database of the meta program.

Meta programming has various application such as ... where thanks to meta
programming it is possible to represent the logic of a language into a formal
and formally verify the wanted properties. On the other hand, it is possible to
embed a logic programming language into another so that some tasks can be
delegated to \ml.

The latter situation motivates our works, since we are implementing a type-class
solver for the \ol \coq in the \ml \elpi (a variant \lambdaprolog). A type class
\cite{wadler89, sozeau08} is a typing structure allowing to introduce ad hoc
polymorphism in functional languages. We call <<instance>> an implementation of
a type class. The resolution of a type-class problem can be viewed as a logic
program where type classes represent predicates parametrized by their arguments
and where instances are rules for those predicates.

@@

Even though type-class resolution is the motivating example of this paper, we
provide a general framework allowing to solve reproduce the same unification
properties of the \ol into the \ml. In other word, if two terms unify in the
\ol, then they still unify in the \ml.

In the following, we consider the \ol being able to quantify over higher-order
variables and accepts $\eta-\beta$-reductions. The same unification properties
are considered valid for the \ml.

There exist two different ways to encode the \ol in the \ml, we can either deep
embed the \ol such that any term of the \ol is represented with a corresponding
predicate. For example, if \texttt{f} is a function of type \coqIn{A -> B} in
the \ol, then the \ml has the predicate \texttt{p} defined as \elpiIn{type p A'
-> B' -> o}, where \elpiIn{A'} and \elpiIn{B'} are types corresponding
respectively to \coqIn{A} and \coqIn{B} in the \ol. In a theorem prover like
\coq, we can translate theorems like the following statement

\begin{equation}
  \texttt{\coqIn{forall F X, p (f X) (fun x => g x (F x)).}}
  \label{eq:coq1}
\end{equation}

where \texttt{p, f} and \texttt{g} are defined constants of the language, into

\begin{equation}
  \texttt{\elpiIn{p (f X) (x\ g x (F x)).}}
\end{equation}

\noindent However, even if this encoding is quite appealing since it allows to
mirror enough straightforwardly the terms of the \ol, we loose the possibility
the manipulate the terms of the \ol into the \ml. In other words, we have no
syntax allowing to know if the current term is a constant, an application, a
lambda abstraction and so on. This is mainly due to the absence of a syntax in
the encoding of the \ol terms. Moreover, another motivation for using syntax to
represent terms of the \ol is that  the typing system of the \ol could
potentially be more expressive than the typing system of the \ml\footnote{This
is the case for \coq wrt \elpi, since in we have no immediate way to encode
the dependent types of \coq into \elpi}.

To simplify the understanding of our encoding, in the following code snippet
we give the typing schema of the \ol terms represented into terms of the \ml.

\begin{mintedTcolorbox}{elpi}
  kind tm type.
  type app list tm -> tm.
  type lam (tm -> tm) -> tm.
  type c string -> tm.
  type uv  nat -> tm.
\end{mintedTcolorbox}

In particular, the type \elpiIn{tm} is the type of the terms of the \ol. The
function applications of the \ol are represented as a list of \elpiIn{tm}
prefixed by the constructor \elpiIn{app}. The \elpiIn{lam} constructor,
represent lambda abstractions of the \ol binding a \elpiIn{tm} into an other
\elpiIn{tm}. Constants as strings inside the constructor \elpiIn{c}. Lastly,
unification variables are integers inside the constructor \elpiIn{uv}, where the
integer is the index of the current variable wrt a list of optional \elpiIn{tm},
standing for the substitution mapping of the \ol.

This second encoding of the \ol into our \ml translate \cref{eq:coq1} into
the term:

\begin{equation}
  \texttt{\elpiIn{app[c "p", app[c "f", uv 0], lam x\ app[c "g", x, app[uv 1, x]]].}}
  \label{eq:elpi1}
\end{equation}

This second encoding of the \ol terms is now structured and as a drawback we are
restricting the unification of the \ol, that is, terms that originally unify
at the \ol level, do not unify in the \ml.

For example, let \coqIn{a} and \coqIn{b} two defined constants and let's try to
unify the \ol term

\begin{equation}
  \texttt{\coqIn{p (f a) (fun x => g x b)}}
\end{equation}

\noindent corresponding to

\begin{equation}
  \texttt{\elpiIn{app[c "p", app[c "f", c "a"], lam x\ app[c "g", x, c "b"]].}}
\end{equation}

\noindent with \cref{eq:coq1} (corresponding to \cref{eq:elpi1}). The
unification of the \ml is able to instantiate \elpiIn{uv 0} (cf \coqIn{X}) to
\elpiIn{c "a"}, but we are no longer capable to unify the sub-term
\elpiIn{app[uv 1, x]} (cf \coqIn{F}) with \elpiIn{c "b"}.

The result of this translation of terms inside the \ml causes a certain lack of
powerfulness while symbolizing higher-order variables. Recall that we are
considering a \ml capable to deal with higher-order variables, however, the
sub-term \elpiIn{app[uv 1, x]} is not expressed into the canonical form where a
higher-order variable of the \ml is in the pattern fragment \cite{miller1991},
i.e. a variable applied to distinct names. Therefore, we need to preprocess the
received unification problem $t_1 = t_2$ by 
\begin{inlinelist}
  \item compiling the terms into a terms $t_1'$ and $t_2'$ understandable by the
  \ml
  \item finding a valid substitution for $t_1'$ and $t_2'$ 
  \item giving back a valid substitution $\theta$ for the \ol, such that $\theta$
  is the most general unifier for $T_1$ and $t_2$ in the logic of the \ol
\end{inlinelist}.

% Let upper-case letter unification variables and lower case variables defined
% constants of the object language. We take as example the following two
% unification problems in our \ol:

% \begin{center}
%   \coqIn{(fun x => P x) = P}
% \end{center}

% The first problem, if

% \begin{center}
%   \elpiIn{p (lam x\ (app F x)) :- ... F ...}
% \end{center}

% cannot unify with goals like

% \begin{center}
%   \elpiIn{p (lam x\ app (app (const "f") x) x).}
% \end{center}

% Since \elpiIn{F} does not see \elpiIn{x} even though it is applied to \elpiIn{x}
% in the object language.

----------------------------- END -----------------------------
\subsection{First-order unification}

@@

\appendix

\begin{table}
  \centering
  \begin{tabular}{c||c|c}
  Term        & \ol                    & \ml                                   \\
  \hline\hline
  Constant    & a                             & `a'                                     \\
  Application & f a\_1 a\_2 ... a\_n          & app{[}`f', `a\_1', `a\_2', ..., `a\_n'{]}     \\
  Abstraction & fun (x : T) $\Rightarrow$ f x & fun `x' T (x\textbackslash app{[}`f', `x'{]}) \\
  Variable    & X                             & X                                             \\
  \end{tabular}
  \caption{\ol terms to \ml terms representation}
  \label{table:coq2elpi}
\end{table}

\section{TC in \coq}
For instance, if
\texttt{XXX} is the type class representing the AAA, then ZZZ and WWW are
instances for XXX. In the code snippet below, we give such implementation in
\coq.

In this paper we do not really want to explain how the translation of the
class/instances is performed in our \ml, we prefer to focus our attention on
unification of terms of the \ol in our \ml. Although, in \cref{table:coq2elpi},
we provide a simple subset of the typing system used to represent the term of
the \ol in the \ml.

Type-class resolution starts from a query, that is a class applied to some
arguments. This \coq term is translated into a term of the \ml and the search
for a solution in the database is started. However, it may happen that the term
representation in the \ml may hide some unification properties that are true in
the \ol. In the example above, the goal \texttt{Decision (@eq T a b)} for some
\texttt{a} and \texttt{b} unifies with \texttt{Decision (R x y)} in the \ol but
not in its meta representation. Similarly, the goal \texttt{RelDecision (@eq
(sig T ?P))} where \texttt{?P}, under the hypothesis \texttt{RelDecision (@eq
nat)}, will trying to apply the rule for \texttt{sig\_eq\_dec}, we fall into an
higher order unification problem, where \texttt{P} is applied to the local name
\texttt{x}. However, the corresponding rule in the \ml exploit a first order
variable \texttt{P}. Therefore, after the refinement of the goal to
\texttt{sig\_eq\_dec}, the resolution immediately fail to solve the premise
\texttt{tc-ProofIrrel (app [P, c0]) (app [P1, c0])}.
